They're Icing up! Team Momentum in Round-Based Esports

### 2022-08-21 paper/paper.Rmd
icing up son

---
title: "They're Clutching up! Team Momentum in Round-Based Esports"
authors:
  - name: Tony ElHabr
    email: anthonyelhabr@gmail.com
    url: https://tonyelhabr.rbind.io
abstract: |
  My research investigates patterns in round win percentages in professional Search and Destroy (SnD) matches of the popular first-person shooter game Call of Duty (CoD). First, I find evidence supprting the hypothesis that round win probability can be modeled as a constant across the series, although not at the naive 50%. Second, I examine post-streak round win probability, given the series outcome. I find no evidence that post-streak win rate is significantly different than expected for combinations of streak length, series length, and series winner, suggesting that the perception of momentum at the team level is just that, perception.
bibliography: references.bib  
output:
  rticles::arxiv_article:
    includes:
      in_header: preamble.tex
---

@@

```{=tex}
\begin{longtable}{crrrrrr}
\caption{Offensive round win rates for the upcoming round, given both the offensive and defensive team's prior number of round wins}\label{tbl:cod-o-win-prop-by-series-state} \\
\toprule
& \multicolumn{6}{c}{Offense round wins} \\ 
\cmidrule(lr){2-7}
Defense round wins & 0 & 1 & 2 & 3 & 4 & 5 \\ 
\midrule
0 & 47.8\%
(852) & 46.6\%
(408) & 43.1\%
(216) & 43.5\%
(115) & 43.3\%
(67) & 40.5\%
(37) \\ 
1 & 48.6\%
(444) & 49.3\%
(418) & 51.5\%
(309) & 43.4\%
(205) & 43.3\%
(120) & 39.4\%
(99) \\ 
2 & 52.8\%
(218) & 48.9\%
(305) & 48.9\%
(315) & 46.6\%
(262) & 48.7\%
(189) & 42.1\%
(133) \\ 
3 & 54.5\%
(123) & 46.0\%
(200) & 49.6\%
(250) & 45.6\%
(248) & 44.4\%
(214) & 44.8\%
(174) \\ 
4 & 56.9\%
(65) & 54.5\%
(145) & 47.2\%
(193) & 44.7\%
(228) & 55.2\%
(221) & 50.5\%
(208) \\ 
5 & 47.4\%
(38) & 49.4\%
(83) & 47.1\%
(136) & 50.9\%
(175) & 45.2\%
(177) & 46.0\%
(202) \\ 
\bottomrule
\end{longtable}
```
# Literature review

@@

Momentum, one of most discussed topics in sports analytics, goes hand-in-hand with a discussion of the nature of series outcomes.[^10] Two opposing fallacies are observed in the context of momentum: the "gambler's fallacy" (negative recency) and the "hot hand fallacy" (positive recency). Per Ayton et al. [-@ayton2004], negative recency is "the belief that, for random events, runs of a particular outcome ... will be balanced by a tendency for the opposite outcome", while positive recency is the expectation of observing future results that match recent results.

[^10]: We often use use "streaks" and momentum interchangeably, but as [@steeger2021] note, momentum implies dependence between events, whereas streaking does not.

Studying both player streaks and team streaks in basketball, in both observational and controlled settings. Gilovich et al. [-@gilovich1985], henceforth GVT, do not find evidence for the hot hand phenomenon. However, Miller and Sanjurjo [-@miller2018], henceforth MS, refuted the conclusions of GVT, proving a framework for quantifying streak selection bias, which effectively works in the manner posited by the gambler's fallacy[^11]. Specifically, MS say that a "bias exists in a common measure of the conditional dependence of present outcomes on streaks of past outcomes in sequential data" implying that, under i.i.d. conditions, "the proportion of successes among the outcomes that immediately follow a streak of consecutive successes is expected to be strictly less than the underlying (conditional) probability of success". When applying streak selection bias to GVT.'s data, MS come to the opposite conclusions as GVT.

[^11]: Streak selection bias, "in conjunction with a quasi-Bayesian model of decision making under sample size neglect... provides a novel structural candidate explanation for the persistence of gambler's fallacy beliefs", per MS.

@@

1.  Both have economic aspects that can create clear advantages for one side in a given round, given how prior rounds played out. CoD has no such equivalent, except for perhaps "kill streaks", which infrequently play a role.
2.  Teams play either offense or defense for many consecutive rounds. (The former has a 14-14-1 format and the latter uses a 12-12-1 format.) On the other hand, teams in CoD SnD rotate roles every round, analogous to a 1-1-1-1-1-1-1 format for home advantage in best-of-seven series for professional sports like the MLB, NBA, and NHL.[^12] While theoretically one might be able to account for any kind of format, such as a 5-5-1, the rotation of team sides every round is convenient for convincing ourselves that rounds could reasonably be modeled as i.i.d. Bernoulli trials.
3.  Both Counter-Strike and Valorant have overtime rules---team must win by two rounds---which can make end-of-series sequences difficult to model. CoD SnD does not have overtime rules.

[^12]: 1-1-1-1-1-1-1 is not used today in these leagues, but it was at least one in each league.

@@

The formula for the expected proportion of series a best-of-$s$ format ending in $r$ rounds ($r \leq s$), $\hat{\Phi}(r)$, for a constant round win probability $\phi$ ($0 \leq \phi \leq 1$) for one team[^13] is

[^13]: If $\phi > 0.5$, then one might say that this team is the better team (known in hindsight).

```{=tex}
\begin{equation}\protect\hypertarget{eq:m}{}{
m = \frac{s + 1}{2}
}\label{eq:m}\end{equation}
```
```{=tex}
\begin{equation}\protect\hypertarget{eq:series-length}{}{
\hat{\Phi}(r) = \frac{(r - 1)!}{(m - 1)!(s - r)!}(\phi^{m}(1 - \phi)^{r - m} + \phi^{r - m}(1 - \phi)^m)
}\label{eq:series-length}\end{equation}
```
$s = 11$ rounds for CoD SnD.[^14]

[^14]: $\phi$ and other symbols are selected so as to reserve symbols like $p$ for usage in other contexts without causing confusion for the reader. Upper-case symbol, e.g. $\Phi$, are consistently used in this paper to represent proportions, while lower-case symbols corresponding to proportions represent probabilities. "Hats", e.g. $\hat{\Phi}$, are used to convey expectations, while "bare" symbols mirroring hat-ed symbols convey observational data.

@@

As shown by MS with their Theorem 1, we should expect the proportion of rounds wins immediately following a streak of $k$ rounds wins for a series lasting $r$ rounds, $\hat{P}^{+|kr}_{MS}$, to be strictly less than $p^{+|r}$.[^15] Unfortunately, there does not exist a closed form representation for the expected value of $\hat{P}^{+|kr}_{MS}$ for $k > 1$. Nonetheless, one may recursively build a collection of dictionaries to compute the expectation analytically, or, more conveniently, one may run simulations to estimate the expected value, as I choose to do.[^16]

[^15]: See Appendix E.1 of MS for the proof.

[^16]: I have adapted the code from Vafa [-@vafa2017], which implements MS's framework [-@miller2018].

Although it would be convenient to emulate the hypothesis testing for difference of proportions performed by MS (and GVT), my context is fundamentally different from that of MS, who focus on longitudinal data in controlled settings.[^17] The number of trials is fixed in their experimental designs, but in CoD SnD, the number of rounds played is determined as a function of the max number of possible rounds ($s$) and whether or not the team wins the series. The Bernoulli trial success probability, i.e. a round win in CoD SnD, is not independent of the opponent. Consequently, a statistical test of the difference in $\hat{P}^{+|kr}$ and $\hat{P}^{-|kr}$, as performed by MS to evaluate their hypothesis regarding post-streak success rates, is not completely appropriate, although useful as a reference.

[^17]: GVT also performs statistical tests on shots from players in live games, i.e. "observational" data, but they note that their findings are likely affected player shot selection in the face of defensive strategy by the opposing team.

@@

We can perform a binomial test to evaluate the null hypothesis $\omega = \omega_0$ for the observed probability of success $\omega$ and a user-specified $\omega_0$, where $0 \leq \omega_0 \leq 1$.[^18] If there are $r^+$ observed successes in a sample of $r$ trials and we expect that there should be $r * \omega_0$, the probability of arriving at this expected number of successes is

[^18]: $\omega$ and other symbols are chosen so as to not conflict with symbols already used elsewhere for in other equations, either generic or for CoD SnD specifically.

@@

Stepping back from the one-sided perspective of a single team's round win probability when streaking, one can attempt to detect the hot hand phenomenon with the Wald-Wolfowitz runs test. Under the null hypothesis, the number of runs in a sequence of $r$ trials is a random variable, which can take on values $r^+$ and $r^-$ for success and failure respectively, with the following mean $\mu$ and variance $\sigma^2$.

```{=tex}
\begin{equation}\protect\hypertarget{eq:ww}{}{
\mu = \frac{2r^{+}r^{-}}{r} + 1, \sigma^2 = \frac{(\mu-1)(\mu-2)}{r-1}
}\label{eq:ww}\end{equation}
```
One might subset observed series sequences to those that violate the run test null hypothesis and perform a test of proportions, where the null is that the observed rate of such series is equal to the frequency that we should expect, given a choice for the constant round win probability $\phi$. If we can reject the null, then we can argue in favor of the hot hand.

@@

First, I investigate the constant probability assumption and the distribution of rounds played in a series. Chance's [-@chance2020] work is closely related to mine, and, in fact, provides a guide for this investigation. Afterwards, I investigate post-streak win rates and the context of momentum, leveraging MS's [-@miller2018] framework for streak selection bias.

@@

Using Equation \ref{eq:chi-squ}, I find that $\chi^2 = 16.0$ (p-value of 0.0068) for the naive constant round win probability $\phi_0 = 0.5$. Thus, I can comfortably reject the constant probability null hypothesis for $\phi_0 = 0.5$, even at a confidence level of $\alpha = 0.01$.

@@

```{=tex}
\begin{longtable}[]{@{}lrr@{}}
\caption{Alternate estimates of the constant probability ($\phi$) for winning a given round in a CoD SnD, applying the three methods suggested by Mosteller (1952), in addition to the naive ($\phi_0 = 0.5$).}\label{tbl:mosteller-methods-results} \\
\toprule()
Method & $\phi$ & $\chi^2$ (p-value) \\
\midrule()
\endfirsthead
\toprule()
Method & $\phi(r)$ & $\chi^2$ (p-value) \\
\midrule()
\endhead
0. Naive & 0.5000 & 16.0 (\textless=0.01) \\
1. Method of moments & 0.5725 & 3.6 (0.6) \\
2. Maximum likelihood & 0.5750 & 3.5 (0.62) \\
3. Minimum ($\chi^2$) & 0.5775 & 3.5 (0.62) \\
\bottomrule()
\end{longtable}
```
Table \ref{tbl:expected-series-lengths-alternative-ps} shows the new $\hat{\Phi}(r)$ when re-applying Equation \ref{eq:series-length} for the maximum likelihood estimate $\phi_2 = 0.575$, resulting in a new set of expected proportions of series lasting $r$ rounds $\hat{\Phi}_2(r)$.[^19] I observe that $\hat{\Phi}_2(r)$ is larger than $\hat{\Phi}_0(r)$ for $r \in [6, 7]$, more closely matching $\Phi(r)$. $\hat{\Phi}_2(r)$ is also closer to the observed $\Phi(r)$ for $r \in [9, 10]$, although not for $r \in [8, 11]$.

[^19]: The method of moments and minimum $\chi^2$ estimates for $\phi$ are omitted simply because the results would be nearly identical to those for the maximum likelihood estimate of $\phi$ (since they are all $\approx 0.575$).

@@

Given that people typically perceive streaks as beginning after the third success (or failure) at minimum [@carlson2007], I focus on streaks of three round wins.[^20] Table \ref{tbl:cod-pw3r-pl3r} compares the observed round win rate $P^{w,kr}$ following streaks of $k=3$ round wins given the series outcome ($r$ rounds and winner) with the notional and MS expected proportions, $\hat{P}^{+|kr}_0$ and $\hat{P}^{+|kr}_{MS}$ respectively.

[^20]: Three happens to also be a reasonable number for series that lasts at maximum 11 rounds.

@@

\bottomrule
\end{longtable}
```
With the exception of $\hat{P}^{+|kr}_{MS}$ when $r = 10$ and the eventual series loser is the team that wins after a streak of three round wins[^21], all un-adjusted binomial test $p$-values are greater than the typical $\alpha = 0.05$ threshold, implying that we cannot reject the null hypothesis that the expected notional and MS post-streak round win rates are different than the observed proportion. Adjusting the $p$-values with the Benjamini & Yekutieli (BY) correction [-@benjamini2001], the null hypothesis cannot be rejected for any case.[^22]

[^21]: This implies that the opponent, the eventual series winner, goes on to win the series 6-4 after going down 0-4. This is the only scenario in which one team wins four rounds consecutively but loses a best-of-11 series in 10 rounds.

[^22]: As noted before, the MS expectations may be unreliable for CoD SnD, so one is inclined to prefer the results of the notional binomial tests.

When performing the same tests for streaks of two, four, and five, there is no case in which I can reject the binomial null hypothesis for the expected notional probability $p^{+|kr}_0$. The null hypothesis can only be rejected for the expected MS probability $p^{+|kr}_{MS}$ when $k = 5, r = 7$ after applying the BY p-value correction.[^23]

[^23]: At 36, the sample size for \$k = 5, r = 7\$ is not as large as most other combinations of \$k\$ and \$r\$.

@@

Now let us explicitly consider the round, $i$. As shown in Table \ref{tbl:cod-pw3ri-pl3ri}, the notional win rate in the round $i$ immediately following a streak of $k$ round wins in a series lasting $r$ rounds, $\hat{P}{+|kri}_0$, is statistically different than the observed $P^{+|kri}$ in several cases, notably when $r = i$ for each of $r \in [9, 10, 11]$.[^24] Interestingly, the observed rate is greater than the expected rate in each case where we can reject the null hypothesis, whereas the observed rate is less than the expected rate in all other cases, as well in a strong majority of cases not shown.

[^24]: The null can also be rejected for $r \in [7, 8]$, although these are not shown, and the sample sizes are smaller.

@@

Further regarding future work, I've reached out to the Twitter user "R11stats", who privately tracks in-round player engagements.[^25] R11stats expressed intent on making the data public, which would allow for research into player-specific momentum.

[^25]: <https://twitter.com/R11stats>

>>>>

---
title: "They're Icing up! Team Momentum in Round-Based Esports"
authors:
  - name: Tony ElHabr
    email: anthonyelhabr@gmail.com
    url: https://tonyelhabr.rbind.io
abstract: |
  My research investigates patterns in round win percentages in professional Search and Destroy (SnD) matches of the popular first-person shooter game Call of Duty (CoD). First, I find evidence supprting the hypothesis that round win probability can be modeled as a constant across the series, although not at the naive 50%. Second, I examine post-streak round win probability, given the series outcome. I find no evidence that post-streak win rate is significantly different than expected for combinations of streak length, series length, and series winner, suggesting that the perception of momentum at the team level is just that, perception.
bibliography: references.bib  
output:
  rticles::arxiv_article:
    includes:
      in_header: preamble.tex
---

@@

```{=tex}
\begin{longtable}{crrrrrr}
\caption{Offensive round win rates for the upcoming round, given both the offensive and defensive team's prior number of round wins}\label{tbl:cod-o-win-prop-by-series-state} \\
\toprule
& \multicolumn{6}{c}{Offense round wins} \\ 
\cmidrule(lr){2-7}
Defense round wins & 0 & 1 & 2 & 3 & 4 & 5 \\ 
\midrule
0 & 47.8\%
(852) & 46.6\%
(408) & 43.1\%
(216) & 43.5\%
(115) & 43.3\%
(67) & 40.5\%
(37) \\ 
1 & 48.6\%
(444) & 49.3\%
(418) & 51.5\%
(309) & 43.4\%
(205) & 43.3\%
(120) & 39.4\%
(99) \\ 
2 & 52.8\%
(218) & 48.9\%
(305) & 48.9\%
(315) & 46.6\%
(262) & 48.7\%
(189) & 42.1\%
(133) \\ 
3 & 54.5\%
(123) & 46.0\%
(200) & 49.6\%
(250) & 45.6\%
(248) & 44.4\%
(214) & 44.8\%
(174) \\ 
4 & 56.9\%
(65) & 54.5\%
(145) & 47.2\%
(193) & 44.7\%
(228) & 55.2\%
(221) & 50.5\%
(208) \\ 
5 & 47.4\%
(38) & 49.4\%
(83) & 47.1\%
(136) & 50.9\%
(175) & 45.2\%
(177) & 46.0\%
(202) \\ 
\bottomrule
\end{longtable}
```
Anecdotally, observers swear by the existence of momentum in CoD SnD series, to the extent that vernacular has been developed to describe such phenomenon. People say that teams "ice up" when the team wins several consecutive rounds to come back to win a series, often culminating with a series-clinching round win.[^10] Further, viewers have come to embrace the "5-3" phenomenon, where teams win three consecutive rounds after facing a 5-3 deficit to win 6-5. There is even a term for the rare 0-5 comeback---a "full sail".

[^10]: This vernacular is not specific to CoD SnD in specific, and is not only used in the contexts of streaks.

# Literature review

@@

Momentum, one of most discussed topics in sports analytics, goes hand-in-hand with a discussion of the nature of series outcomes.[^11] Two opposing fallacies are observed in the context of momentum: the "gambler's fallacy" (negative recency) and the "hot hand fallacy" (positive recency). Per Ayton et al. [-@ayton2004], negative recency is "the belief that, for random events, runs of a particular outcome ... will be balanced by a tendency for the opposite outcome", while positive recency is the expectation of observing future results that match recent results.

[^11]: We often use use "streaks" and momentum interchangeably, but as [@steeger2021] note, momentum implies dependence between events, whereas streaking does not.

Studying both player streaks and team streaks in basketball, in both observational and controlled settings. Gilovich et al. [-@gilovich1985], henceforth GVT, do not find evidence for the hot hand phenomenon. However, Miller and Sanjurjo [-@miller2018], henceforth MS, refuted the conclusions of GVT, proving a framework for quantifying streak selection bias, which effectively works in the manner posited by the gambler's fallacy[^12]. Specifically, MS say that a "bias exists in a common measure of the conditional dependence of present outcomes on streaks of past outcomes in sequential data" implying that, under i.i.d. conditions, "the proportion of successes among the outcomes that immediately follow a streak of consecutive successes is expected to be strictly less than the underlying (conditional) probability of success". When applying streak selection bias to GVT.'s data, MS come to the opposite conclusions as GVT.

[^12]: Streak selection bias, "in conjunction with a quasi-Bayesian model of decision making under sample size neglect... provides a novel structural candidate explanation for the persistence of gambler's fallacy beliefs", per MS.

@@

1.  Both have economic aspects that can create clear advantages for one side in a given round, given how prior rounds played out. CoD has no such equivalent, except for perhaps "kill streaks", which infrequently play a role.
2.  Teams play either offense or defense for many consecutive rounds. (The former has a 14-14-1 format and the latter uses a 12-12-1 format.) On the other hand, teams in CoD SnD rotate roles every round, analogous to a 1-1-1-1-1-1-1 format for home advantage in best-of-seven series for professional sports like the MLB, NBA, and NHL.[^13] While theoretically one might be able to account for any kind of format, such as a 5-5-1, the rotation of team sides every round is convenient for convincing ourselves that rounds could reasonably be modeled as i.i.d. Bernoulli trials.
3.  Both Counter-Strike and Valorant have overtime rules---team must win by two rounds---which can make end-of-series sequences difficult to model. CoD SnD does not have overtime rules.

[^13]: 1-1-1-1-1-1-1 is not used today in these leagues, but it was at least one in each league.

@@

The formula for the expected proportion of series a best-of-$s$ format ending in $r$ rounds ($r \leq s$), $\hat{\Phi}(r)$, for a constant round win probability $\phi$ ($0 \leq \phi \leq 1$) for one team[^14] is

[^14]: If $\phi > 0.5$, then one might say that this team is the better team (known in hindsight).

```{=tex}
\begin{equation}\protect\hypertarget{eq:m}{}{
m = \frac{s + 1}{2}
}\label{eq:m}\end{equation}
```
```{=tex}
\begin{equation}\protect\hypertarget{eq:series-length}{}{
\hat{\Phi}(r) = \frac{(r - 1)!}{(m - 1)!(s - r)!}(\phi^{m}(1 - \phi)^{r - m} + \phi^{r - m}(1 - \phi)^m)
}\label{eq:series-length}\end{equation}
```
$s = 11$ rounds for CoD SnD.[^15]

[^15]: $\phi$ and other symbols are selected so as to reserve symbols like $p$ for usage in other contexts without causing confusion for the reader. Upper-case symbol, e.g. $\Phi$, are consistently used in this paper to represent proportions, while lower-case symbols corresponding to proportions represent probabilities. "Hats", e.g. $\hat{\Phi}$, are used to convey expectations, while "bare" symbols mirroring hat-ed symbols convey observational data.

@@

As shown by MS with their Theorem 1, we should expect the proportion of rounds wins immediately following a streak of $k$ rounds wins for a series lasting $r$ rounds, $\hat{P}^{+|kr}_{MS}$, to be strictly less than $p^{+|r}$.[^16] Unfortunately, there does not exist a closed form representation for the expected value of $\hat{P}^{+|kr}_{MS}$ for $k > 1$. Nonetheless, one may recursively build a collection of dictionaries to compute the expectation analytically, or, more conveniently, one may run simulations to estimate the expected value, as I choose to do.[^17]

[^16]: See Appendix E.1 of MS for the proof.

[^17]: I have adapted the code from Vafa [-@vafa2017], which implements MS's framework [-@miller2018].

Although it would be convenient to emulate the hypothesis testing for difference of proportions performed by MS (and GVT), my context is fundamentally different from that of MS, who focus on longitudinal data in controlled settings.[^18] The number of trials is fixed in their experimental designs, but in CoD SnD, the number of rounds played is determined as a function of the max number of possible rounds ($s$) and whether or not the team wins the series. The Bernoulli trial success probability, i.e. a round win in CoD SnD, is not independent of the opponent. Consequently, a statistical test of the difference in $\hat{P}^{+|kr}$ and $\hat{P}^{-|kr}$, as performed by MS to evaluate their hypothesis regarding post-streak success rates, is not completely appropriate, although useful as a reference.

[^18]: GVT also performs statistical tests on shots from players in live games, i.e. "observational" data, but they note that their findings are likely affected player shot selection in the face of defensive strategy by the opposing team.

@@

We can perform a binomial test to evaluate the null hypothesis $\omega = \omega_0$ for the observed probability of success $\omega$ and a user-specified $\omega_0$, where $0 \leq \omega_0 \leq 1$.[^19] If there are $r^+$ observed successes in a sample of $r$ trials and we expect that there should be $r * \omega_0$, the probability of arriving at this expected number of successes is

[^19]: $\omega$ and other symbols are chosen so as to not conflict with symbols already used elsewhere for in other equations, either generic or for CoD SnD specifically.

@@

Stepping back from the one-sided perspective of a single team's round win probability when streaking, one can attempt to detect the hot hand phenomenon with the Wald-Wolfowitz runs test. Under the null hypothesis, the number of runs in a sequence of $r$ trials, $\zeta(r)$, is a random variable that can take on values $r^+$ and $r^-$ for success and failure respectively, with the following mean $\mu$ and variance $\sigma^2$.

```{=tex}
\begin{equation}\protect\hypertarget{eq:ww}{}{
\mu = \frac{2r^{+}r^{-}}{r} + 1, \sigma^2 = \frac{(\mu-1)(\mu-2)}{r-1}
}\label{eq:ww}\end{equation}
```
To incorporate my findings regarding constant round win probability, I can specify that $\Pr(r^+) = \phi$ (and, conversely, that $\Pr(r^-) = 1 - \phi$).

One can subset the observed series sequences to those that violate the run test null hypothesis and perform a test of proportions, where the null is that the observed rate $\psi^+$ of such series is equal to the frequency rate $\psi_0$. The test statistic $Z$

```{=tex}
\begin{equation}\protect\hypertarget{eq:prop}{}{
Z = \frac{\psi^+ - \psi^+_0}{\sqrt{\psi^+_d (1 - \psi^+_d) ( \frac{1}{N^+} + \frac{1}{\hat{N}^+})}}
}\label{eq:prop}\end{equation}
```
where $\psi^+_d = \frac{r^+ - \hat{r}^+}{N^+ - \hat{N}^+}$. If we can reject the null hypothesis for many sequences, then there is implicit evidence in favor of the hot hand.

@@

First, I investigate the constant probability assumption and the distribution of rounds played in a series. Chance's [-@chance2020] work is closely related to mine, and, in fact, provides a guide for this investigation. Afterwards, I investigate post-streak win rates and the context of momentum.

@@

Using Equation \ref{eq:chi-squ}, I find that $\chi^2 = 16.0$ ($p$-value of 0.0068) for the naive constant round win probability $\phi_0 = 0.5$. Thus, I can comfortably reject the constant probability hypothesis for the null $\phi_0 = 0.5$, even at a confidence level of $\alpha = 0.01$.

@@

```{=tex}
\begin{longtable}[]{@{}lrr@{}}
\caption{Alternate estimates of the constant probability ($\phi$) for winning a given round in a CoD SnD, applying the three methods suggested by Mosteller (1952), in addition to the naive ($\phi_0 = 0.5$).}\label{tbl:mosteller-methods-results} \\
\toprule()
Method & $\phi$ & $\chi^2$ (p-value) \\
\midrule()
\endfirsthead
\toprule()
Method & $\phi(r)$ & $\chi^2$ (p-value) \\
\midrule()
\endhead
0. Naive & 0.5000 & 16.0 (\textless=0.01) \\
1. Method of moments & 0.5725 & 3.6 (0.6) \\
2. Maximum likelihood & 0.5750 & 3.5 (0.62) \\
3. Minimum ($\chi^2$) & 0.5775 & 3.5 (0.62) \\
\bottomrule()
\end{longtable}
```
Table \ref{tbl:expected-series-lengths-alternative-ps} shows the new $\hat{\Phi}(r)$ when re-applying Equation \ref{eq:series-length} for the maximum likelihood estimate $\phi_2 = 0.575$, resulting in a new set of expected proportions of series lasting $r$ rounds $\hat{\Phi}_2(r)$.[^20] I observe that $\hat{\Phi}_2(r)$ is larger than $\hat{\Phi}_0(r)$ for $r \in [6, 7]$, more closely matching $\Phi(r)$. $\hat{\Phi}_2(r)$ is also closer to the observed $\Phi(r)$ for $r \in [9, 10]$, although not for $r \in [8, 11]$.

[^20]: The method of moments and minimum $\chi^2$ estimates for $\phi$ are omitted simply because the results would be nearly identical to those for the maximum likelihood estimate of $\phi$ (since they are all $\approx 0.575$).

@@

Given that people typically perceive streaks as beginning after the third success (or failure) at minimum [@carlson2007], I focus on streaks of three round wins.[^21] Table \ref{tbl:cod-pw3r-pl3r} compares the observed round win rate $P^{w,kr}$ following streaks of $k=3$ round wins given the series outcome ($r$ rounds and winner) with the notional and MS expected proportions, $\hat{P}^{+|kr}_0$ and $\hat{P}^{+|kr}_{MS}$ respectively.

[^21]: Three happens to also be a reasonable number for series that lasts at maximum 11 rounds.

@@

\bottomrule
\end{longtable}
```
With the exception of $\hat{P}^{+|kr}_{MS}$ when $r = 10$ and the eventual series loser is the team that wins after a streak of three round wins[^22], all un-adjusted binomial test $p$-values are greater than the typical $\alpha = 0.05$ threshold, implying that we cannot reject the null hypothesis that the expected notional and MS post-streak round win rates are different than the observed proportion. Adjusting the $p$-values with the Benjamini & Yekutieli (BY) correction [-@benjamini2001], the null hypothesis cannot be rejected for any case.[^23]

[^22]: This implies that the opponent, the eventual series winner, goes on to win the series 6-4 after going down 0-4. This is the only scenario in which one team wins four rounds consecutively but loses a best-of-11 series in 10 rounds.

[^23]: As noted before, the MS expectations may be unreliable for CoD SnD, so one is inclined to prefer the results of the notional binomial tests.

When performing the same tests for streaks of two, four, and five, there is no case in which I can reject the binomial null hypothesis for the expected notional probability $p^{+|kr}_0$. The null hypothesis can only be rejected for the expected MS probability $p^{+|kr}_{MS}$ when $k = 5, r = 7$ after applying the BY p-value correction.[^24]

[^24]: At 36, the sample size for \$k = 5, r = 7\$ is not as large as most other combinations of \$k\$ and \$r\$.

@@

Now let us explicitly consider the round, $i$. As shown in Table \ref{tbl:cod-pw3ri-pl3ri}, the notional win rate in the round $i$ immediately following a streak of $k$ round wins in a series lasting $r$ rounds, $\hat{P}{+|kri}_0$, is statistically different than the observed $P^{+|kri}$ in several cases, notably when $r = i$ for each of $r \in [9, 10, 11]$.[^25] Interestingly, the observed rate is greater than the expected rate in each case where we can reject the null hypothesis, whereas the observed rate is less than the expected rate in all other cases, as well in a strong majority of cases not shown.

[^25]: The null can also be rejected for $r \in [7, 8]$, although these are not shown, and the sample sizes are smaller.

@@

Further regarding future work, I've reached out to the Twitter user "R11stats", who privately tracks in-round player engagements.[^26] R11stats expressed intent on making the data public, which would allow for research into player-specific momentum.

[^26]: <https://twitter.com/R11stats>

### 2022-08-21 paper/paper.tex
icing up son

\title{They're Clutching up! Team Momentum in Round-Based Esports}

@@

@@

Stepping back from the one-sided perspective of a single team's round
win probability when streaking, one can attempt to detect the hot hand
phenomenon with the Wald-Wolfowitz runs test. Under the null hypothesis,
the number of runs in a sequence of \(r\) trials is a random variable,
which can take on values \(r^+\) and \(r^-\) for success and failure
respectively, with the following mean \(\mu\) and variance \(\sigma^2\).

@@

One might subset observed series sequences to those that violate the run
test null hypothesis and perform a test of proportions, where the null
is that the observed rate of such series is equal to the frequency that
we should expect, given a choice for the constant round win probability
\(\phi\). If we can reject the null, then we can argue in favor of the
hot hand.

@@

First, I investigate the constant probability assumption and the
distribution of rounds played in a series. Chance's (2020) work is
closely related to mine, and, in fact, provides a guide for this
investigation. Afterwards, I investigate post-streak win rates and the
context of momentum, leveraging MS's (2018) framework for streak
selection bias.

@@

Using Equation \ref{eq:chi-squ}, I find that \(\chi^2 = 16.0\) (p-value
of 0.0068) for the naive constant round win probability
\(\phi_0 = 0.5\). Thus, I can comfortably reject the constant
probability null hypothesis for \(\phi_0 = 0.5\), even at a confidence
level of \(\alpha = 0.01\).

>>>>

\title{They're Icing up! Team Momentum in Round-Based Esports}

@@

Anecdotally, observers swear by the existence of momentum in CoD SnD
series, to the extent that vernacular has been developed to describe
such phenomenon. People say that teams ``ice up'' when the team wins
several consecutive rounds to come back to win a series, often
culminating with a series-clinching round win.\footnote{This vernacular
  is not specific to CoD SnD in specific, and is not only used in the
  contexts of streaks.} Further, viewers have come to embrace the
``5-3'' phenomenon, where teams win three consecutive rounds after
facing a 5-3 deficit to win 6-5. There is even a term for the rare 0-5
comeback---a ``full sail''.

@@

Stepping back from the one-sided perspective of a single team's round
win probability when streaking, one can attempt to detect the hot hand
phenomenon with the Wald-Wolfowitz runs test. Under the null hypothesis,
the number of runs in a sequence of \(r\) trials, \(\zeta(r)\), is a
random variable that can take on values \(r^+\) and \(r^-\) for success
and failure respectively, with the following mean \(\mu\) and variance
\(\sigma^2\).

@@

To incorporate my findings regarding constant round win probability, I
can specify that \(\Pr(r^+) = \phi\) (and, conversely, that
\(\Pr(r^-) = 1 - \phi\)).

One can subset the observed series sequences to those that violate the
run test null hypothesis and perform a test of proportions, where the
null is that the observed rate \(\psi^+\) of such series is equal to the
frequency rate \(\psi_0\). The test statistic \(Z\)

\begin{equation}\protect\hypertarget{eq:prop}{}{
Z = \frac{\psi^+ - \psi^+_0}{\sqrt{\psi^+_d (1 - \psi^+_d) ( \frac{1}{N^+} + \frac{1}{\hat{N}^+})}}
}\label{eq:prop}\end{equation}

where \(\psi^+_d = \frac{r^+ - \hat{r}^+}{N^+ - \hat{N}^+}\). If we can
reject the null hypothesis for many sequences, then there is implicit
evidence in favor of the hot hand.

@@

First, I investigate the constant probability assumption and the
distribution of rounds played in a series. Chance's (2020) work is
closely related to mine, and, in fact, provides a guide for this
investigation. Afterwards, I investigate post-streak win rates and the
context of momentum.

@@

Using Equation \ref{eq:chi-squ}, I find that \(\chi^2 = 16.0\)
(\(p\)-value of 0.0068) for the naive constant round win probability
\(\phi_0 = 0.5\). Thus, I can comfortably reject the constant
probability hypothesis for the null \(\phi_0 = 0.5\), even at a
confidence level of \(\alpha = 0.01\).
